---
title: "Stats202C-HW2"
author: "Juan Piao"
date: "5/22/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem1.
### Calculate the mean of a Gamma(4.3, 6.2) random variable using (a) Accept-Reject with a Gamma(4, 7) candidate. (b) Metropolis- Hastings with a Gamma(4, 7) candidate. (c) Metropolis- Hastings with a Gamma(5, 6) candidate.

```{r}
set.seed(123456)
b<-4.3/4
#M<- b^(-4)*((4.3-4)/((1-b)*exp(1)))^(4.3-4) 

M <- gamma(4) / gamma(4.3) * b**4 * (4.3 * exp(-1))**(4.3 - 4) * 6.2^4.3/7^4

g47=rgamma(5000,4,1/7)
u=runif(5000,max=dgamma(g47,4,1/7))
x=g47[u<dgamma(g47,4.3,1/6.2)/M]

ylim <- c(0, 0.04)

par(mfrow=c(1,3),mar=c(4,4,1,1))
hist(x,freq=FALSE,xlab="",ylab="",col="wheat2", ylim = ylim,
main="a: Accept-Rej with Ga(4.7) prop")
curve( dgamma(x,4.3,1/6.2),lwd=2,col="sienna",add=T)
a <- length(x)/5000
mean(x)


###### Met-Hast with Ga(5,6) prop
X=rep(0,5000)

X[1]=rgamma(1,4.3,1/6.2)

for (t in 2:5000){
  rho=(dgamma(X[t-1],4,1/7)*dgamma(g47[t],4.3,1/6.2))/(dgamma(g47[t],4,7)*dgamma(X[t-1],4.3,1/6.2))

  X[t]=X[t-1]+(g47[t]-X[t-1])*(runif(1)<rho) #move to g47[t] if it is accepted, otherwise stay
}
hist(X,freq=FALSE,xlab="",ylab="",col="wheat2",  ylim = ylim,
main="b: Met-Hast with Ga(4,7) prop")
curve(dgamma(x,4.3,1/6.2),lwd=2,col="sienna",add=T)

b <- length(unique(X))/5000
mean(X)

########### Met-Hast with Ga(5,6) prop
X2=rep(0,5000)

g56=rgamma(5000,5,1/6)
X2[1]=rgamma(1,4.3,1/6.2)
for (t in 2:5000){
rho=(dgamma(X2[t-1],5,1/6)*dgamma(g56[t],4.3,1/6.2))/
(dgamma(g56[t],5,1/6)*dgamma(X2[t-1],4.3,1/6.2))
X2[t]=X2[t-1]+(g56[t]-X2[t-1])*(runif(1)<rho)
}
c <- length(unique(X2))/5000
hist(X2,freq=FALSE,xlab="",ylab="",col="wheat2",  ylim = ylim,
main="c: Met-Hast with Ga(5,6) prop")
curve(dgamma(x,4.3,1/6.2),lwd=2,col="sienna",add=T)
mean(X2)

```

```{r}

plot(cumsum(x)/1:length(x), type = "l", ylim = c(10, 60), ylab = "mean")
lines(cumsum(X)/1:length(X), col="red")
lines(cumsum(X2)/1:length(X2), col="blue")
legend("topright", legend=c("Accept-Reject with Gamma(4, 7)", "Metropolis-Hastings with Gamma(4, 7)",  "Metropolis-Hastings with Gamma(5, 6)"), col=c("black", "red", "blue"), lty=1)
```


The mean of a Gamma(4.3, 6.2) random variable using 
(a) Accept-Reject with a Gamma(4, 7) candidate is 26.61735.
(b) Metropolis-Hastings with a Gamma(4, 7) candidate is 27.8449.
(c) Metropolis-Hastings with a Gamma(5, 6) candidate is 26.59489.


## Problem 7.20 

### (a) Verify the likelihood function

$logit(p_{ij})=a+bx_i+cz_{ij}$, $i=1,...,k$, $j=1,...,n_i$
Solving for $p_{ij}$, we get, 
\begin{align*}
p_{ij}&= \frac{exp(a+bx_i+cz_{ij})}{1+exp(a+bx_i+cz_{ij})}\\
\text{The likelihood function is, }\\
&\prod_{i=1}^{k} \prod_{j=1}^{n_i} (p_{ij})^{y_{ij}}(1-p_{ij})^{1-y_{ij}}\\
=&\prod_{i=1}^{k} \prod_{j=1}^{n_i} (\frac{exp(a+bx_i+cz_{ij})}{1+exp(a+bx_i+cz_{ij})})^{y_{ij}}(1-\frac{exp(a+bx_i+cz_{ij})}{1+exp(a+bx_i+cz_{ij})})^{1-y_{ij}}\\
=&\prod_{i=1}^{k} \prod_{j=1}^{n_i} (\frac{exp(a+bx_i+cz_{ij})}{1+exp(a+bx_i+cz_{ij})})^{y_{ij}}(\frac{1}{1+exp(a+bx_i+cz_{ij})})^{1-y_{ij}}\end{align*}

### (b) Run a standard GLM on these data9 and get the estimated mean and variance of a, b, and c.

```{r}
LogisticData<-read.table("~/Downloads/Stats 202C/HW2/LogisticData.txt", header = T)

Logisticmodel<-glm(LogisticData$erodd~ LogisticData$np + LogisticData$metq, family = "binomial")
summary(Logisticmodel)
```

The mean and variance of a are -1.97391, 0.0489 respectively.
The mean and variance of b are  0.16221, 0.00636 respectively.
The mean and variance of c are  0.28438, 0.0086 respectively.

### (c) Use normal candidate densities with mean and variance at the GLM estimates in a Metropolis-Hastings algorithm that samples from the likelihood. Get histograms of the parameter values.

```{r}
np = LogisticData$np
metq = LogisticData$metq
erodd = LogisticData$erodd

am = -1.97391 # coef (mean)
as = 0.22113 # std err
bm = 0.16221
bs = 0.07974
cm = 0.28438
cs = 0.09274

log_likelihood<- function(a, b, c){
  sum(erodd * (a + b*np + c*metq) - log(1 + exp(a + b*np + c*metq)))
}

get_dens <- function(a, b, c){
  dnorm(a, am, as, log = T) + dnorm(b, bm, bs, log = T)+ dnorm(c, cm, cs, log = T)
}

n_sim = 100000
a_hat <- rep(0, n_sim)
b_hat <- rep(0, n_sim)
c_hat <- rep(0, n_sim)
a_hat[1] <- rnorm(1, am, as)
b_hat[1] <- rnorm(1, bm, bs)
c_hat[1] <- rnorm(1, cm, cs)

for (i in 2: n_sim){
  a = a_hat[i-1]
  b = b_hat[i-1]
  c = c_hat[i-1]
  
  a_tr <- rnorm(1, am, as)
  b_tr <- rnorm(1, bm, bs)
  c_tr <- rnorm(1, cm, cs)

  tr_logl = log_likelihood(a_tr, b_tr, c_tr)
  logl = log_likelihood(a, b, c)
  A = tr_logl - logl
  
  dens  = get_dens(a, b, c)
  tr_dens = get_dens(a_tr, b_tr, c_tr)
  B = dens - tr_dens

  P = exp(A + B)
  rho = runif(1) < min(P, 1)
  a_hat[i] = a_tr*rho + a_hat[i-1]*(1-rho)
  b_hat[i] = b_tr*rho + b_hat[i-1]*(1-rho)
  c_hat[i] = c_tr*rho + c_hat[i-1]*(1-rho)
  
}

```

```{r}
print( paste0("a_mean: ", round( mean(a_hat), 4)))
print( paste0("b_mean: ", round( mean(b_hat), 4)))
print( paste0("c_mean: ", round( mean(c_hat), 4)))

print( paste0("a_variance: ", round( var(a_hat), 4)))
print( paste0("b_variance: ", round( var(b_hat), 4)))
print( paste0("c_variance: ", round( var(c_hat), 4)))
```


### parameter a histogram and density estimate

```{r}
hist(a_hat, freq = F, main ='Histogram of Parameter a (Intercept)', col = "wheat", breaks = 50)
#curve(dnorm(x,mean(a_hat), sd(a_hat)),lwd=2,col="sienna",add=T)
 abline(v=am,lwd=2,col=2)
lines(density(a_hat), col="sienna", lwd=2)
legend("topright", legend = c("Density", "true value"), col=c("sienna", "red"), lty=1, lwd = 2)

```







### parameter a histogram and density estimate

```{r}
hist(b_hat, freq = F, main ='Histogram of Parameter b (for HMO Type)', col="wheat", breaks = 50)
#curve(dnorm(x,mean(b_hat), sd(b_hat)),lwd=2,col="sienna",add=T)
 abline(v=bm,lwd=2,col=2)
lines(density(b_hat), col="sienna", lwd=2)
legend("topright", legend = c("Density", "true value"), col=c("sienna", "red"), lty=1, lwd = 2)

```

```{r}
hist(c_hat, freq = F, main ='Histogram of Parameter c (for Health Status)', col="wheat", breaks = 50)
#curve(dnorm(x,mean(c_hat), sd(c_hat)),lwd=2,col="sienna",add=T)
 abline(v=cm,lwd=2,col=2)
lines(density(c_hat), col="sienna", lwd=2)
legend("topright", legend = c("Density", "true value"), col=c("sienna", "red"), lty=1, lwd = 2)

```

```{r, include=FALSE}
### Compare the result with "MCMClogit" 

library("MCMCpack")
library("lattice")
model.mcmc <- MCMClogit(erodd~ np + metq, data = LogisticData)
xyplot(model.mcmc)
par(mfrow=c(1,3),mar=c(4,4,1,1))

```


```{r, include=FALSE}
par(mfrow=c(1,3),mar=c(4,4,1,1))

densplot(model.mcmc)
```

